---
title: distRcpp
output: md_document
---

```{r, include = FALSE}
options(width = 100)
```

![](https://img.shields.io/badge/dev-beta-red.svg)
[![GitHub release](https://img.shields.io/github/release/btskinner/distRcpp.svg)](https://github.com/btskinner/distRcpp)
[![R build status](https://github.com/btskinner/distRcpp/workflows/R-CMD-check/badge.svg)](https://github.com/btskinner/distRcpp/actions)

This package uses [Rcpp](http://www.rcpp.org) to quickly compute population/distance-weighted measures. Geodesic distances can be computed using either [Haversine](https://en.wikipedia.org/wiki/Haversine_formula) or [Vincenty](https://en.wikipedia.org/wiki/Vincenty%27s_formulae) formulas. The package also has functions to return raw distance measures. If you are able to [install Rcpp on your machine](https://github.com/RcppCore/Rcpp), you should be able to install this package and use these functions.

Install the latest development version from Github with

```{r, eval = FALSE}
devtools::install_github("btskinner/distRcpp")
```

**NB** This package is still in early beta stages. It does not have much in the way of error handling. Data must be pre-processed so that no missing (`NA`) values are given to the functions.

## Available functions

### Weighted measures

#### `dist_weighted_mean()`

Interpolate values for a vector of locations (**X**) that are the inverse-distance-weighted average of measures taken at surrounding locations (**Y**). For each point, *x*, nearby values of the measure taken at **Y** are weighted more heavily than those from locations that are farther away.

#### `popdist_weighted_mean()`

Interpolate values for a vector of locations (**X**) that are the population/inverse-distance-weighted average of measures taken at surrounding locations (**Y**). For each point, *x*, nearby values of the measure taken at **Y** are weighted more heavily than those from locations that are farther away. Measures taken in more heavily populated *y* are given more weight than those with lower populations. This weighting scheme is a compromise between distance and population and is useful for interpolating measures that need to take both into account.

### Distances

#### `dist_1to1()`

Compute and return the geodesic distance between two spatial points. Returns distance in meters.

#### `dist_1tom()`

Compute and return the geodesic distance between one location and a vector of other locations. Returns vector of distances in meters.

#### `dist_mtom()`

Compute and return the geodesic distance between each coordinate pair in two vectors. Returns *n x k* matrix of distances in meters, where *n* = # of locations in first vector and *k* = # of locations in second vector.

#### `dist_df()`

Compute distance between corresponding coordinate pairs and return vector of distances in meters. For use when creating a new `data.frame` or [dplyr](https://CRAN.R-project.org/package=dplyr) `tbl_df()` column.

#### `dist_min()`

Compute minimum distance between each starting point, *x*, and
possible end points, **Y**. Returns vector of minimum distances in
meters that equals # of starting points (size of **X**).

#### `dist_max()`

Compute maximum distance between each starting point, *x*, and possible end points, **Y**. Returns vector of maximum distances in meters that equals # of starting points (size of **X**).

## Benchmark

Compare speed with base R function when measuring the distance between every United States population-weighted county centroid as measured in 2010 (N = 3,143 with complete measurements).

### Load data

```{r, results = 'hide', message = FALSE}
## libraries
libs <- c("dplyr","microbenchmark","geosphere","distRcpp")
lapply(libs, require, character.only = TRUE)
```
```{r}
## read data
df <- get(data(county_centers))
df

## subset to 2010 population-weighted centroids (pclon10, pclat10)
p <- df %>% select(pclon10, pclat10) %>% na.omit %>% data.frame()

```

### Check for equality
```{r}
dist_R <- distm(p)
dist_Rcpp <- dist_mtom(p[,1],p[,2],p[,1],p[,2])

dist_R[1:5,1:5]
dist_Rcpp[1:5,1:5]
all.equal(dist_R, dist_Rcpp)
```

### Benchmark

2018 MacBookPro, 2.9 GHz Intel Core i9, 32 GB 2400 MHz DDR4 SDRAM

```{r}
microbenchmark(
    dist_R = distm(p),
    dist_Rcpp = dist_mtom(p[,1],p[,2],p[,1],p[,2]),
    times = 100
)
```

### Big file
```{r}
## get census block group centers of population
bg <- readr::read_csv("https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG.txt") %>%
    setNames(tolower(names(.))) %>%
    filter(statefp < 56) %>%
    mutate(id = paste0(statefp, countyfp, tractce, blkgrpce),
           lon = longitude,
           lat = latitude) %>%
    select(id, lon, lat) %>%
    na.omit()

ct <- get(data(county_centers)) %>%
    rename(id = fips,
           lon = pclon10,
           lat = pclat10) %>%
    na.omit()
bg
ct

time(dist_Rcpp <- dist_min(x_df = ct, y_df = bg))

```
