---
title: distRcpp
output: md_document
---

```{r, include = FALSE}
options(width = 100)
```

![](https://img.shields.io/badge/dev-beta-red.svg)
[![GitHub release](https://img.shields.io/github/release/btskinner/distRcpp.svg)](https://github.com/btskinner/distRcpp)
[![R build status](https://github.com/btskinner/distRcpp/actions/workflows/check-standard.yml/badge.svg)](https://github.com/btskinner/distRcpp/actions)

This package uses [Rcpp](http://www.rcpp.org) to quickly compute 
population/distance-weighted measures. Geodesic distances can be computed using 
either [Haversine](https://en.wikipedia.org/wiki/Haversine_formula) or 
[Vincenty](https://en.wikipedia.org/wiki/Vincenty%27s_formulae) formulas. The 
package also has functions to return raw distance measures. If you are able to 
[install Rcpp on your machine](https://github.com/RcppCore/Rcpp), you should be 
able to install this package and use these functions.

Install the latest development version from Github with

```{r, eval = FALSE}
pak::pak("btskinner/distRcpp")
```

**NB** This package is still in beta stages. It does not have much in the 
way of error handling. Data must be pre-processed so that no missing (`NA`) 
values are given to the functions.

## Available functions

### Weighted measures

#### `dist_weighted_mean()`

Interpolate values for a vector of locations (**X**) that are the 
inverse-distance-weighted average of measures taken at surrounding locations 
(**Y**). For each point, *x*, nearby values of the measure taken at **Y** are 
weighted more heavily than those from locations that are farther away.

#### `popdist_weighted_mean()`

Interpolate values for a vector of locations (**X**) that are the 
population/inverse-distance-weighted average of measures taken at surrounding 
locations (**Y**). For each point, *x*, nearby values of the measure taken at 
**Y** are weighted more heavily than those from locations that are farther away. 
Measures taken in more heavily populated *y* are given more weight than those 
with lower populations. This weighting scheme is a compromise between distance 
and population and is useful for interpolating measures that need to take both 
into account.

### Distances

#### `dist_1to1()`

Compute and return the geodesic distance between two spatial points. Returns 
distance in meters.

#### `dist_1tom()`

Compute and return the geodesic distance between one location and a vector of 
other locations. Returns vector of distances in meters.

#### `dist_mtom()`

Compute and return the geodesic distance between each coordinate pair in two 
vectors. Returns *n x k* matrix of distances in meters, where *n* = # of 
locations in first vector and *k* = # of locations in second vector.

#### `dist_df()`

Compute distance between corresponding coordinate pairs and return vector of 
distances in meters. For use when creating a new `data.frame` or 
[dplyr](https://CRAN.R-project.org/package=dplyr) `tbl_df()` column.

#### `dist_min()`

Compute minimum distance between each starting point, *x*, and
possible end points, **Y**. Returns vector of minimum distances in
meters that equals # of starting points (size of **X**).

#### `dist_max()`

Compute maximum distance between each starting point, *x*, and possible end 
points, **Y**. Returns vector of maximum distances in meters that equals # of 
starting points (size of **X**).

#### `dist_nearest_n()`

Compute the distance between each starting point, *x*, and possible end points,
**Y** and return the ids and distances to the *n* nearest points (default = 10).

## Benchmark

Compare speed with base R function when measuring the distance between every 
United States population-weighted county centroid as measured in 2010 
(N = 3,143 with complete measurements).

### Load data
```{r, include = FALSE}
options(tidyverse.quiet = TRUE)
```

```{r, results = 'hide', message = FALSE}
## libraries
libs <- c("tidyverse","microbenchmark","geosphere","distRcpp")
sapply(libs, require, character.only = TRUE)
```
```{r}
## read data
df <- get(data(countycentroids))
df

## subset to 2010 population-weighted centroids (pclon10, pclat10)
p <- df |> select(pclon10, pclat10) |> drop_na() |> data.frame()

```

### Check for equality
```{r}
dist_R <- matrix(NA_real_, 5, 5)
px <- p[1:5, ]
py <- p[6:10,]
for (i in 1:nrow(px)) {
    for (j in 1:nrow(py)) {
        dist_R[i,j] <- geosphere::distHaversine(px[i,], py[j,],
            r = 6371008.7714150594547
        )
    }
}
dist_Rcpp <- distRcpp::dist_mtom(px[,1],px[,2],py[,1],py[,2])

dist_R
dist_Rcpp
all.equal(dist_R, dist_Rcpp)
```

### Benchmark

2024 MacBookPro, Apple M4 Pro, 48 GB Memory

```{r}
x <- 1:1000
y <- 1001:2000
microbenchmark(
    dist_R = geosphere::distm(p[x,], p[y,], fun = distHaversine),
    dist_Rcpp = distRcpp::dist_mtom(p[x,1],p[x,2],p[y,1],p[y,2]),
    times = 100
)
```

### Big file
```{r}
## get census block group centers of population
file_url <- file.path(
    "https://www2.census.gov/geo/docs/reference",
    "cenpop2010/blkgrp/CenPop2010_Mean_BG.txt"
)
bg <- readr::read_csv(file_url, show_col_types = FALSE) |> 
    rename_all(tolower) |> 
    filter(statefp < 56) |> 
    mutate(id = paste0(statefp, countyfp, tractce, blkgrpce),
           lon = longitude,
           lat = latitude) |> 
    select(id, lon, lat) |> 
    drop_na()

ct <- get(data(countycentroids)) |> 
    rename(id = fips,
           lon = pclon10,
           lat = pclat10) |> 
    drop_na()
bg
ct

system.time(dist_Rcpp <- distRcpp::dist_min(x_df = ct, y_df = bg))
dist_Rcpp |> tibble()
```
